BOLD Parameter Estimation Using Particle Filters

### 2010-06-29 doc/thesis/background.tex
done with properties I think

\chapter{Current Techniques}
The ultimate purpose of this work is to provide a new set of tools
for analyzing FMRI data. Whereas esisting techniques have been 
successful at finding macroscopic regions of activation, :hese
techniques are not robust to significant noise and thus linear 
modeling carries a significant bias error due to lack of model
flexibility. While adding parameters can significantly increase
error due to model variance, this effect should be mitigated first
by the use of a highly constrained model based on first principals
and also by the
calculation of a full posterior distribution, rather than a single
estimate. The
purpose of this paper is thus to evaluate the potential of using
a particle filter along with the BOLD model to derive physical 
parameters. In so doing, we hope to be able to show that one or more
parameters are a suitable replacement for estimating voxel 
activation from a standard FMRI image. We also hope to show that 
estimated posterior distribution of the parameters, derived from
the particle filter, is able to provide an accurate measure of the
confidence interval.

\section{Previous Studies of Parameters}
There have been quite a few efforts to quantify the parameters of the
various BOLD models.  Although \cite{Buxton1998} and \cite{Friston2000}
both proposed physiologically reasonable values for the model parameters, 
\cite{Friston2002} was the first paper to calculate the parameters based 
on actual FMRI data. In that paper, Friston et. al. used a variation of
Expectation Maximization to find a normal distribution for the parameters:
\begin{eqnarray}
\epsilon &=& N(.54 , .1 ^2 )  \nonumber \\
\tau_s & =&  N(1.54, .25^2)   \nonumber \\
\tau_f & =&  N(2.46, .25^2)   \nonumber \\
\tau_0 & =&  N(.98 , .25^2 )   \nonumber \\
\alpha & =&  N(.33 , .45^2 )   \nonumber \\
E_0   & =&  N(.34 ,  .1 ^2 )   \nonumber \\
V_0  & = &  .03 (not\ estimated) \nonumber
\end{eqnarray}
Since then, several other methods of have been used to calculate
BOLD parameters from FMRI timecourses. In \cite{Riera2004}, a maximum
likelihood method for innovation processes was used, as described by
\cite{Ozaki1994}. \cite{Ozaki1994} uses a similar construction to a 
Kalman filter, to break the time series into a series of innovations,
for which Maximum Likelihood was performed. While this is in some since the
"right" way to find the solution, it comes with several caveats. First, every
step in parameter space requires a recalculation of all the state variables. With
two or three parameters this is fine, more than that, and calculations could go on
indefinitely. Second, it still assumes the parameters and noise are Gaussian, and
will only be optimal in that case. Third, depending on the nonlinearities present
in the system, local minima may be extremely common. Later \cite{Hu2009} used an 
Unscented Kalman Filter over all the parameters and state variables to find the 
parameter set/variable time series. While this method has the drawback of not necessarily
being optimal, unfortunately there is no general optimal solution to non-linear non
Gaussian models. Hu et. al.'s technique also will run significantly faster than
ML based techniques, since it does not require recalculating the entire timeseries
for every step in parameters. Both \cite{Hu2009} and \cite{Friston2002} came to results 
very similar to the expected values stated in \cite{Buxton1998} and \cite{Friston2000}.
One potential problem with all these techniques are that the depend heavily on the
priors. The starting point of the parameters could have a huge impact on the
results, and while one can be hopeful that this isn't the reason for the agreement
between \cite{Friston2000} and later results, there is no way to know.

In \cite{Johnston2008}, a hybrid particle filter/gradient
descent algorithm was used to simultaneously derive the static (classicly called
parameters) and dynamic parameters (classically known as state variables).
Essentially a particle filter is used to calculate the state variables at some
time, and then the estimated distribution of the particles was used to find
the most likely set of parameters that would give that distribution of state variables.
\cite{Johnston2008} comes to a very different set of parameter estimates as compared
to the original \cite{Friston2000} guesses:
\begin{eqnarray}
epsilon &=& .069 \pm .014    \nonumber \\
tau_s & =& 4.98 \pm 1.07  \nonumber \\
tau_f & =& 8.31 \pm 1.51   \nonumber \\
tau_0 & =& 8.38 \pm 1.5    \nonumber \\
alpha & =& .189 \pm .004   \nonumber \\
E_0   & =& .635 \pm .072     \nonumber \\
V_0   & =& .0149 \pm .006     \nonumber 
\end{eqnarray}
Notably, the time constants are significantly longer. This could be a result of
some preprocessing to the stimulus timeseries performed in \cite{Friston2002} and
later works but not in \cite{Johnston2008}, or it could be that \cite{Johnston2008}
depended less on the priors.

@@

\section{Basic Statistical Parametric Mapping}
Although not strictly the same thing as parameter calculation from 
FMRI, activation detection is very similar. In fact, estimation of 
parameters is somewhat a generalization of the idea of activation detection.
Therefore, it is important to draw a comparison between the methods proposed
in this thesis with existing methods of activation detection.

The most basic method of analyzing FMRI data is through a standard T-test
between "resting state" and "active state" samples. This is done by 
taking the average and variance of the inactive period, and the 
period during which the stimulus was activate separately then treating 
them both as gaussian distributions.
If they are in fact Gaussian distributions, then a basic t-test will
give the probability that the samples came from the same distribution
(the null hypothesis). Of course, this test is fraught with problems; even if
the drift mentioned earlier has been removed, there is little reason
to believe that the noise is Gaussian, or even stable. Additionally, 
even if the noise were Gaussian, a t-test with a p-value of .05 over
50000 or more samples is on average going to generate $.05*50000$ false
positives. To compensate for this, bonferoni correction, also known as
multiple comparison tests are performed; essentially p-values are 
divided by the number of independent
tests being run. This, however, leads to extremely low p-values, so
low that it would be impossible for any biological system to satisfy. To
compensate, a Gaussian kernel is applied to the image, thus reducing
variance (and thus separating the active and inactive distributions)
as well as decreasing the effective number of voxels. Since t-tests are
now no longer being applied to n <I need to define n> independent voxels,
the factor by which the p-value must be divided by can be decreased.
<Do I need to mathematically define all this?> The derivation and application
of random field theory, and its use can be found in various papers \cite{Worsley2004}.

\subsection{General Linear Model}
\label{sec:Current Techniques General Linear Model}
The most used form of FMRI analysis is Statistical Parametric
Mapping, but is able to account for several different levels or types
of stimulus (see \cite{Hofmann1997}). By using hierarchical models
the output signal timeseries is considered the weighted
sum of the various input timeseries. Essentially every experimental factor
is considered as another level inputs (ex. different type of stimulus,
different run with the same patient, different patient). 
The equation for a general linear model is then
\begin{equation}
Y(t) = X(t)\beta + \epsilon(t)
\end{equation}
where $Y(t)$ is the smoothed or detrended timeseries of measurements,
$X(t)$ is a row vector of input, $\beta$ is a column vector of weights,
and $\epsilon$ is the error. Thus for every time, the measurement is
assumed to be a weighted sum of the columns of $X$ plus some error. The calculation
of $\beta$ is then performed using a maximum likelihood or gradient descent search 
to minimize the error.

\begin{figure}
\caption{GLM todo}
\label{fig:GLM}
\end{figure}

It is well known of course that a square wave stimulus does not result in a square wave
in the activation of brain regions. Thus, various methods are used to 
smooth the columns of $X$, and thus bandlimit the input. 
The best technique is convolving the stimulus input with a Hemodynamic 
Response Function (HRF), which mimics the basic shape of BOLD activation, including a delay
due to rise time and fall time. The downside of this method, however is that 
the shape of the Hemodynamic Signal is static, meaning the same Hemodynamic Function is
used for every region of the brain, and $Y(t)$ must linear combination 
of scalar the columns of $X$ to be appropriately identified. Additionally, 
it is well known that different Hemodynamic Response Functions are necessary for different 
regions of the brain. The "Canonical" HRF that is most used, has been fitted
for the visual cortex. Thus the shape of the HRF may differ significantly in
terms of onset and fall time in other areas of the brain \cite{Handwerker2004}. 
The inability to
fit parameters other than scale certainly hinders SPM's
ability to locate voxels that do not conform to the "canonical" hemodynamic 
response function. If there are different HRF's for different regions 
of the brain, might there also be more subtle differences even within
a particular region? It seems reasonable to think so. As a result, the 
most common use of SPM will be heavily biased toward the visual cortex.

The GLM is extremely powerful at determining the linear dependence of
a set of regressors on the output. Unfortunately, there is significant evidence
that many of these dependencies are nonlinear, which means they may be
difficult or impossible to detect using linear techniques. This presents
a significant and unknown problem that is often left un-addressed in 
neural studies.

A static linear model is also unable to incorporate other forms of physiological
data that may be available. Combined FMRI
CBF or CBV imaging methods are not uncommon and they could shed much light on
neural activation. However, there is no way of actively combining that data 
into a static HRF. The fact of the matter is that the relation between the 
BOLD signal and CBV/CBF simply cannot be described by any linear relationship.
While it may be possible to determine some regionally varying set of parameters
for the HRF based on CBV/CBF, the lack of physiologically inspired parameters
impairs this. The importance of regional differences in the HRF cannot be
over-stated; it is well known that capillary beds are far from uniformly distributed
and thus blood perfusion and regional oxygenation cannot possibly be uniform
either.

Finally, all these techniques require the noise to be Gaussian to reach
an optimal solution. In fact there is no known optimal solution to
nonlinear models with non-Guassian noise, so obviously some assumptions
are going to have to be made to reach a solution. However, it would be
nice to have an algorithm that is robust to these effects, and could
still give a good solution when Gaussianity is violated.

Because of these limitations, its entirely possible that activation exists
in regions that SPM doesn't detect, but because the activation does not conform
to the set HRF, it is impossible for these signals to be detected. Perhaps
because of this, it is not uncommon for data to be thrown out or averaged
together in FMRI studies because 
no significant activation could be seen in a single run \cite{Riera2004}
\cite{Johnston2008}. This practice highlights 
limitations in strictly linear approaches; and suggests that a single HRF is 
insufficiently flexible to account for relatively common variations of neural activation
\cite{Handwerker2004}.

In general activation detection type methods also don't have the ability 
to find pathologies based on state variables or parameters.  It
is quite possible that physical properties such as decreased compliance of
blood vessels could indicate a neurological condition that is not easily
seen in a T1 or T2 map. In essence, this could make FMRI a much more 
useful clinical tool than it is now.

>>>>

\chapter{Prior Works}
The ultimate purpose of this work is to provide a new tool
for analyzing FMRI data.

Currently, FMRI is used to determine the location of responses
to stimuli. The method of finding activation is discussed in 
\autoref{sec:Statistical Parameteric Mapping}. The goal of this 
work is to move away from the question "Is this region active"
and instead ask "What does the activation look like". Answering
this question necessitates more complex models and certainly
will result in longer run times. Already there have been several other
attempts to model the BOLD response; these works will be discussed
in [all other sections basically].

\section{Statistical Parametric Mapping}
Although not strictly the same as parameter calculation from 
FMRI, activation detection is similar and worth discussing. Estimation of 
parameters is a generalization of the idea of activation detection.
Given the popularity of Statistical Parametric Mapping (SPM) 
it is important to draw a distinction between the methods proposed
in this work.

\subsection{Square Wave Model}
\label{sec:Square Wave}
The most basic method of analyzing FMRI data is through a standard T-test
between "resting state" and "active state" samples. Simply put, the
mean is calculated separately for non-stimulus and stimulus time intervals.
A classic t-test may then be applied, giving the probability that the
distributions actually have the same mean. Because of the correlated
noise present in FMRI (\autoref{sec:Introduction Noise}), it is necessary
to apply some sort of high-pass filter to the data. Without applying
such a filter, P values must be set extraordinarily high to prevent
false positives \cite{Smith2007}. If there truly is signal
due to stimuli, the distributions will not actually be independent
Gaussians because activation does not fit a square wave 
(\autoref{sec:BOLD Physiology}). For this reason other methods
are often more used, as discussed in 
\autoref{sec:Current Techniques General Linear Model}.

\subsection{Random Field Theory}
SPM methods make significant use of T-Tests across large regions;
however, such T-tests work slightly differently than a single
individual test. A t-test with a p-value of $.05$ over a modestly sized FMRI image,
say 10,000 voxels, will on average generate 500 false
positives. This is called the multiple comparison problem. Traditional
Bonferroni Correction
deals with this by requiring each test to pass with P
value of $\frac{.05}{10000}$. The probability of 
a single false positive would then $.05$. Unfortunately this leads to unrealistically
low p-values; so low that it would be impossible for any biological system to satisfy. To
compensate, a Gaussian kernel is applied to smooth the image. This has
the benefit of reducing the noise variance and 
decreasing the effective number of independent measurements. Because
the number of independent measurements is smaller, Bonferroni correction
can theoretically be applied with a lower scaling factor than
the original voxel count \cite{Worsley2004}. A side effect of this,
a single voxel activation is virtually impossible to detect.

\subsection{General Linear Model}
\label{sec:Current Techniques General Linear Model}
The most common FMRI analysis technique is SPM, though there
are more advanced versions that the simple square wave
method discussed in \autoref{sec:Square Wave}. Hierarchal
Models are one important improvement that allows researchers
to combine data across multiple runs, patients and stimuli
(see \cite{Hofmann1997} for more on Hierarchical Modeling). 
Hierarchical Models
concatenate all the data into a single dataset, then perform
a linear fit between a design matrix and the data. The design matrix
encapsulates all known experimental factors such as stimuli,
young/old, etc.  The equation for a general linear model is:

\begin{equation}
Y(t) = X(t)\beta + \epsilon(t)
\end{equation}

where $Y(t)$ is the smoothed or de-trended time course of measurements,
$X(t)$ is the design matrix, $\beta$ is a column vector of weights,
and $\epsilon$ is the error. Thus for every time, the measurement is
assumed to be a weighted sum of the columns of $X$ plus some error. The calculation
of $\beta$ is then performed using a maximum likelihood or gradient descent search 
to minimize the error.

As mentioned previously, a square wave stimulus 
does not result in a square wave in the activation of brain regions. 
The BOLD signal is in fact a significantly smoothed version of the 
stimuli. As such, when fitting an FMRI time course to the input,
the input ($X$'s columns) is usually smoothed to reduce bias error. 
The best method, that maintains a linear fit, is convolving the input 
with a Hemodynamic 
Response Function (HRF). The so called "Hemodynamic Response Function"
mimics the basic shape of BOLD activation, including a delay
due to rise time and fall time. The fitting 
process is then a least squares fit over the space of the vector $\beta$. 
Therefore $Y(t)$ is estimated as a linear combination 
of the columns of $X$.

Smoothing the input with a single HRF poses certain problems.
It is well known that different Hemodynamic Response Functions are necessary 
for different regions of the brain. The "Canonical" HRF that is most used, 
has been optimized for the visual cortex. As \autoref{sec:Post Stimulus Undershoot}
discussed, there are certainly variations in the shape of the
BOLD response, both between brain regions and patients.  \cite{Handwerker2004}
discusses the implications of choosing an incorrect HRF, the most important
of which is a definite increase in false negatives. While an atlas
of Hemodynamic Response Functions for each region could definitely 
mitigate this risk, it does not deal with variation between patients.
Thus, the inability to fit parameters other than scale definitely hinders 
analysis of activation. As a result, there is significant bias error
when using a linear fit to localize activation. Notably, studies
will also have a bias toward the visual cortex, precisely because
it is so well studied.

\subsection{Hierarchical Linear Models}
As mentioned previously and discussed extensively
in \cite{Friston2002} and \cite{Hoffman1997}, hierarchical models may be 
applied to account for systematic differences between subjects. For instance, if the
study happens to be a mix between young and old, incorporating
that into the model is wise, regardless if that is the purpose of the test. 
The reason to do this is to account for additional variance that may not
be present in similar studies. The Hierarchical form used by
\cite{Friston2002} is shown in \autoref{eq:Hiearchical}.

\begin{eqnarray}
\label{eq:Hierarchical}
Y(t) = X_1(t)\theta_1 + \epsilon_1(t)            \nonumber \\
\theta_1(t) = X_2(t)\theta_2 + \epsilon_2(t)     \nonumber \\
...                                              \nonumber \\
\theta_{n-1}(t) = X_n(t)\theta_n + \epsilon_n(t) 
\end{eqnarray}

The Empirical Bayes algorithm is used in both in both \cite{Friston2002} 
and \cite{Hoffman1997}. As a consequence, point estimators 
are used for each $\theta$, rather than the full distributions.

\subsection{Conclusion}
In all, the GLM is extremely useful for determining linear 
dependence of a set of regressors on the output. Unfortunately, as discussed in
\autoref{sec:BOLD Analysis} there are significant nonlinearities 
that almost certainly cause false negatives in the Statistical Parametric
Maps. Unfortunately nonlinear analyses have only recently become feasible,
so the scale of the problem is still unknown. The problem is 
highlighted by the relatively common scenario where no significant
activation can be found in a single run \cite{Riera2004}
\cite{Johnston2008}.

The static nature of the  linear model also limits its inference power. 
Besides not permitting HRF differences between patients, there is no
reasonable way to incorporate other forms of physiological
data. Combined FMRI CBF or CBV imaging methods are rapidly getting better,
as seen in \cite{Chan2009}. These techniques could shed light on
neural activation by providing extra measurements, yet a 
physiologically reasonable model is necessary for this.
In reverse, activation detection methods also don't have the ability 
to identify pathologies based on state variables or parameters. For
example decreased compliance of
blood vessels could indicate, or even cause, a neurological condition that 
is not easily seen in other imaging modalities. Thus, the benefits
of physiologically meaningful models are manifold.

\section{Previous Studies of the BOLD Model}
Unlike Statistical Parametric Mapping, the techniques described in this
section are all attempts to learn some version of the BOLD model. 
Of particular interest are the calculated distributions for the 
BOLD parameters, since they are used as prior distributions for
the particles filter, described in \autoref{sec:Methods}.

%todo
There have been quite a few efforts to quantify the parameters of the
various BOLD models.  Although \cite{Buxton1998} and \cite{Friston2000}
both proposed physiologically reasonable values for the model parameters, 
\cite{Friston2002} was the first paper to calculate the parameters based 
on actual FMRI data. In that paper, Friston et. al. used a variation of
Expectation Maximization to find a normal distribution for the parameters:
Since then, several other methods of have been used to calculate
BOLD parameters from FMRI timecourses. In \cite{Riera2004}, a maximum
likelihood method for innovation processes was used, as described by
\cite{Ozaki1994}. \cite{Ozaki1994} uses a similar construction to a 
Kalman filter, to break the time series into a series of innovations,
for which Maximum Likelihood was performed. While this is in some since the
"right" way to find the solution, it comes with several caveats. First, every
step in parameter space requires a recalculation of all the state variables. With
two or three parameters this is fine, more than that, and calculations could go on
indefinitely. Second, it still assumes the parameters and noise are Gaussian, and
will only be optimal in that case. Third, depending on the nonlinearities present
in the system, local minima may be extremely common. Later \cite{Hu2009} used an 
Unscented Kalman Filter over all the parameters and state variables to find the 
parameter set/variable time series. While this method has the drawback of not necessarily
being optimal, unfortunately there is no general optimal solution to non-linear non
Gaussian models. Hu et. al.'s technique also will run significantly faster than
ML based techniques, since it does not require recalculating the entire timeseries
for every step in parameters. Both \cite{Hu2009} and \cite{Friston2002} came to results 
very similar to the expected values stated in \cite{Buxton1998} and \cite{Friston2000}.
One potential problem with all these techniques are that the depend heavily on the
priors. The starting point of the parameters could have a huge impact on the
results, and while one can be hopeful that this isn't the reason for the agreement
between \cite{Friston2000} and later results, there is no way to know.

In \cite{Johnston2008}, a hybrid particle filter/gradient
descent algorithm was used to simultaneously derive the static (classicly called
parameters) and dynamic parameters (classically known as state variables).
Essentially a particle filter is used to calculate the state variables at some
time, and then the estimated distribution of the particles was used to find
the most likely set of parameters that would give that distribution of state variables.
\cite{Johnston2008} comes to a very different set of parameter estimates as compared
to the original \cite{Friston2000} guesses:
Notably, the time constants are significantly longer. This could be a result of
some preprocessing to the stimulus timeseries performed in \cite{Friston2002} and
later works but not in \cite{Johnston2008}, or it could be that \cite{Johnston2008}
depended less on the priors.

@@
